Effectful Semantics for Substructural SSA

### 2023-11-15 pldi24/paper.tex
foo
starting text for related work

Our grammar proper is split into three main syntactic categories:
\begin{itemize}
  \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the expression \(a\)
  is of type \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or \(p = 0\)
  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). The rules
  for this judgement are given in Figure \ref{fig:ssa-term-typing}.
  \item \textit{Blocks} \(s, t\), which are typed via the judgement
  \(\haslb{\ms{L}}{t}{\Gamma}\), which says that the block \(t\) is well-formed
  with respect to the context \(\Gamma\) and branches only to labels in
  \(\ms{L}\). The rules for this judgement are given in Figure
  \ref{fig:ssa-block-typing}.
  \item \textit{Label-sets} or \textit{control-flow graphs} \(L\), which are
  typed via the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\)
  defines a control-flow graph with inputs \(\ms{L}\) and outputs \(\ms{K}\)
  (some of which may also be in \(\ms{L}\), indicating recursive control-flow).
  The rules for this judgement are given in Figure \ref{fig:ssa-block-typing}.
\end{itemize}
Here \textit{contexts} \(\Gamma\), consist of a list of variables \(x\) given a
type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\). A variable
\(x: A^q\) is called \textit{affine} if its type \(A\) is affine (i.e.
\(\tlin{A}{\taff}\)) \textit{and} \(\taff \subseteq q\), in which case it can be
freely ignored. Similarly, \(x\) is \textit{relevant} if its type \(A\) relevant
(i.e. \(\tlin{A}{\trel}\)) \textit{and} \(\trel \subseteq q\), in which case it
can be freely ignored. This is formalized in the structural judgement
\(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be split into
subcontexts \(\Delta\) and \(\Xi\) (weakening is simply the special case
\(\cwk{\Gamma}{\Delta} \iff \csplits{\Gamma}{\Delta}{\cdot}\)), as per the
typing rules in Figure \ref{fig:ssa-structural}. In particular, the rule
\rle{split-nil} provides a base case by saying that the empty context can be
split into two empty contexts, while the rules \rle{split-left} and
\rle{split-right} send \(x: A^q\) to the left or right subcontext repsectively;
both these rules also allow us to reduce the quantity \(q\) to a quantity \(r
\subseteq q\), "forgetting" some of our capabilities. The rule \rle{split-dup}
allows the duplication of a variable which has a relevant type \(A\)
\textit{and} a relevant quantity \(q\), with the option to weaken the quantity
to \textit{different} quantities \(r\) and \(s\) on the left-hand side and
right-hand sides respectively. Finally, the rule \rle{split-aff} simply allows
us to ignore a variable \(x: A^q\) if \(A\) and \(q\) are affine.

@@

@@

\paragraph{Type Theory for SSA}

\TODO{\citet{hua-explicit-ssa-2010}}
\TODO{\citet{ssa-types-matsuno-ohori-06}}
\TODO{\citet{typed-effect-ssa-rigon-torrens-vasconcellos-20}}

\paragraph{Functional SSA}

\TODO{\citet{chakravarty-functional-ssa-2003}}

\subparagraph{Operational Semantics}

\TODO{\citet{barthe-compcert-ssa-2014}}
\TODO{\citet{buchwald-verified-ssa-2016}}
\TODO{\citet{herklotz-gsa-2023}}
\TODO{\citet{garbuzov-structural-cfg-2018}}
\TODO{\citet{vellum}}

>>>>

Our grammar proper is split into three main syntactic categories:
\begin{itemize}
  \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the expression \(a\)
  is of type \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or \(p = 0\)
  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). The rules
  for this judgement are given in Figure \ref{fig:ssa-term-typing}.

  So the centrality $p$ indicates whether it is safe to substitute this expression
  for a variable, and $q$, plus the linearity of $A$, control how many times that
  variable can occur.

  \item \textit{Blocks} \(s, t\), which are typed via the judgement
  \(\haslb{\ms{L}}{t}{\Gamma}\), which says that the block \(t\) is well-formed
  with respect to the context \(\Gamma\) and branches only to labels in
  \(\ms{L}\). The rules for this judgement are given in Figure
  \ref{fig:ssa-block-typing}.

\item \textit{Label-sets} or \textit{control-flow graphs} \(L\), which are
  typed via the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\)
  defines a control-flow graph with inputs \(\ms{L}\) and outputs \(\ms{K}\)
  (some of which may also be in \(\ms{L}\), indicating recursive control-flow).
  The rules for this judgement are given in Figure \ref{fig:ssa-block-typing}.
\end{itemize}
Here \textit{contexts} \(\Gamma\), consist of a list of variables \(x\) given a
type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\). A variable
\(x: A^q\) is called \textit{affine} if its type \(A\) is affine (i.e.
\(\tlin{A}{\taff}\)) \textit{and} \(\taff \subseteq q\), in which case it can be
freely ignored. Similarly, \(x\) is \textit{relevant} if its type \(A\) relevant
(i.e. \(\tlin{A}{\trel}\)) \textit{and} \(\trel \subseteq q\), in which case it
can be freely ignored. This is formalized in the structural judgement
\(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be split into
subcontexts \(\Delta\) and \(\Xi\) (weakening is simply the special case
\(\cwk{\Gamma}{\Delta} \iff \csplits{\Gamma}{\Delta}{\cdot}\)), as per the
typing rules in Figure \ref{fig:ssa-structural}. In particular, the rule
\rle{split-nil} provides a base case by saying that the empty context can be
split into two empty contexts, while the rules \rle{split-left} and
\rle{split-right} send \(x: A^q\) to the left or right subcontext repsectively;
both these rules also allow us to reduce the quantity \(q\) to a quantity \(r
\subseteq q\), "forgetting" some of our capabilities. The rule \rle{split-dup}
allows the duplication of a variable which has a relevant type \(A\)
\textit{and} a relevant quantity \(q\), with the option to weaken the quantity
to \textit{different} quantities \(r\) and \(s\) on the left-hand side and
right-hand sides respectively. Finally, the rule \rle{split-aff} simply allows
us to ignore a variable \(x: A^q\) if \(A\) and \(q\) are affine.

@@

Originally, \citet{kelsey-ssa-cps} showed how to embed SSA into a CPS
representation, and \citet{appel-ssa} showed informally how SSA can be
seen as a collection of mutually-tail-recursive
procedures. \citet{chakravarty-functional-ssa-2003} made Appel's
observation formal, by showing how to translate SSA into A-normal
form~\cite{anf}, and showed how to use the translation to prove the
correctness of a constant propagation analysis. These translations
were for an untyped source and target language, and
\citet{typed-effect-ssa-rigon-torrens-vasconcellos-20} give a typed
translation from SSA into the lambda calculus, which also tracks the
effects of expressions. The effect tracking is handled Koka-style,
where the effect system tracks a list of effect names. (In contrast,
we track the centrality and duplicability of terms, without reference
to the specific effect.)

All of these papers give SSA semantics by translation -- that is,
they defined the semantics of an SSA program to be the semantics of
the image of the translation, whether CPS or ANF. There have also
been a number of papers which have studied the semantics of SSA directly.

\citet{ssa-types-matsuno-ohori-06} give a type theory for what appear
to be ordinary three-address code programs, where the types of the
variables at each program point represent the control flows from the
assignment statements reaching that program point. However, the type
information means that every well-typed program can be placed into
SSA-form by inserting $\phi$-nodes in a completely type-directed
way. This lets them avoid having to give an operational reading of
$\phi$-nodes, since they can use the standard semantics for
three-address code.

\citet{hua-explicit-ssa-2010} give another type system for SSA
programs, as well as a direct operational semantics for it, using the
standard sequential heap semantics, as well as proving progress and
preservation for their type system.

There have also been many operational semantics for SSA arising from
compiler verification efforts. \citet{barthe-compcert-ssa-2014} give
a

\TODO{}}
\TODO{\citet{buchwald-verified-ssa-2016}}
\TODO{\citet{herklotz-gsa-2023}}
\TODO{\citet{garbuzov-structural-cfg-2018}}
\TODO{\citet{vellum}}

@@
